Epidemiological birth-death models with partner notification

### 2024-09-02 article/article_plos.tex
Rewrote the PN extension justifications

We developed an extension of phylodynamic multi-type birth-death (MTBD) models with partner notification (PN), and a simulator to generate trees under MTBD and MTBD-PN models. We proposed a non-parametric test for detecting partner notification in pathogen phylogenetic trees. Its application to simulated data showed that it is both highly specific and sensitive. For the simplest representative of the MTBD-PN family, the BD-PN model, we developed a maximum likelihood parameter and CI estimator. It performed accurate estimations on simulated data, and detected partner notification in HIV-1 B epidemics in Zurich and the UK. Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Phylodynamic models can estimate epidemiological parameters from pathogen phylogenetic trees (i.e., genealogies). These models typically assume that sampling of pathogen sequences from detected cases is independent between infected individuals. However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification (contact tracing) is included in health policies of many countries.

We developed a phylodynamic model accounting for partner notification and a test for detection of partner notification in pathogen phylogenetic trees. They showed good performance both on simulated and real data (HIV-1 B epidemics in Zurich and the UK). Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation, while our new estimator performs well both in presence and in absence of partner notification.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e., genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g., due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BD model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g., SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g., Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g., HIV) more than the others~\cite{Stadler2014}.

@@

Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling. One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (contact tracing). Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection. It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}. For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.

@@

In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN. We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models. We propose a non-parametric test for detecting partner notification in pathogen phylogenetic trees. For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator. We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemics in Zurich and in the UK. Importantly, we show that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.

@@

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum likelihood framework). In particular, it permits inference of the following epidemiological parameters:

@@

We propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$. The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to health-care providers for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g., got diagnosed though the standard procedure), they get tested upon notification. The parameter $\upsilon$ hence regroups a complex set of conditions, which would hardly be identifiable separately from only sequence data. Upon notification, the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$). Upon sampling, the partner might notify in their turn. BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate. This creates an observed tip in the transmission tree. 
\end{itemize}

@@

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in~\nameref{S1_Appendix}. The results in section~"\nameref{sec:test:sim}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.

@@

For each model, we generated 100 trees with 500--1\,000 tips. For each tree the parameter values were drawn randomly within the following bounds:
\begin{itemize}
\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
\item sampling probability $\rho \in ]0.1, 0.9]$,
\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
\end{itemize}

The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models.

@@

Let us assume that there are N cherries in the tree. For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$. We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $2i*$ gets swapped with the tip $2i-1*$ ($1 \leq i \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}. (If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle as in Fig.~\ref{fig:tipswap}). We then calculate the time differences $d_{i'}$ between the tips of these shuffles cherries. 
The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to month or even a year). Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries. However, if some cherries are ``notified'' their real tip differences should be smaller than reshuffled ones. We therefore output the pvalue calculated by the sign test.

@@

We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its true value (for identifiability). %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material. 
As expected with a maximum likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.

>>>>

We developed an extension of phylodynamic multi-type birth-death (MTBD) models with partner notification (PN), and a simulator to generate trees under MTBD and MTBD-PN models. We proposed a non-parametric test for detecting partner notification in pathogen phylogenetic trees. Its application to simulated data showed that it is both highly specific and sensitive. For the simplest representative of the MTBD-PN family, the BD-PN model, we solved the differential equations and proposed a closed form solution for the likelihood function, which we implemented in a software program to estimate the model parameters and their confidence intervals from phylogenetic trees. It performed accurate estimations on simulated data, and detected partner notification in HIV-1 B epidemics in Zurich and the UK. Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Phylodynamic models can estimate epidemiological parameters (e.g., the number of secondary infections $R_e$) from pathogen phylogenetic trees (i.e., genealogies, estimated from sampled pathogen genomic sequences). These models typically assume that sampling of pathogen sequences from detected cases is independent between infected individuals. However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification (contact tracing) is included in health policies of many countries.

We developed a phylodynamic model accounting for partner notification and a test for the detection of partner notification in pathogen phylogenetic trees. Our test and epidemiological parameter estimator showed good performance both on simulated and real data. We detected the presence of partner notification in the HIV-1 B epidemics in Zurich and the UK, and corrected the previous parameter estimates made without accounting for partner notification. %  Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation, while our new estimator performs well both in presence and in absence of partner notification.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e., genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}. Using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however, for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g., due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

However, the spread and detection or an epidemic can be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extension of the classical BD model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}. The MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). MTBD models are phylodynamic analogies of compartmental models in classical epidemiology (e.g., SIR, Susceptible-Infectious-Removed).  Examples of MTBD models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g., Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g., HIV) more than the others~\cite{Stadler2014}.

@@

Another important source of heterogeneity that none of these models accounts for is caused by non-independent sampling. One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (PN). PN (or contact tracing) is a process in which contacts of detected infected individuals are identified and offered testing for infection. It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}. For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.

@@

In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification, MTBD-PN. We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models. We propose a non-parametric test for detecting partner notification in pathogen phylogenetic trees. For the simplest representative of this model family, the BD-PN model, we also derive the equations (and their closed-form solution) for its likelihood calculation and present a maximum-likelihood parameter estimator. We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemics in Zurich and in the UK. Importantly, we show that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.

@@

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum-likelihood framework). In particular, it permits the estimation of the following epidemiological parameters:

@@

We propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  
In the partner notification process, an identified infected index case notifies their potentially infected contacts (e.g., sexual or drug injection partners for HIV), either on their own (patient referral) or with a hep of a health service provider (provider referral, when a health adviser notifies the partners without identifying the index patient; or contract referral, when if the index case fails to notify partners within a specified time period, the health adviser proceeds to provider referral)~\cite{wardPartnerNotification2014}. The notified partners are then provided with appropriate tests and treatment. Therefore, a successful detection of a new case B via partner notification by an index case A implies several conditions to be met: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (ii) B is successfully reached and notified; (iii) B agrees to take a test; (iv) B is indeed infected. 
For instance, according to the HIV report from the Public Health England~\cite{aHIVUnitedKingdom2014}, in England in 2013, reaching the stage (ii) had an estimated probability of about 50\%; reaching the stage (iii) -- of about 41\% (as 83\% of all the patients who attended a clinic through PN had a reported HIV test); and the stage (iv) -- of less than 3\%.

All these conditions are hardly identifiable separately from only sequence data, hence in our model we combine them into one, ``an infected person B being successfully notified by A'', and model it with a constant notification probability $\upsilon$ which is applied upon sampling of A.

For simplicity, a positive draw of $\upsilon$ models a slightly modified scenario: (i) A remembers the contact with B and  either notifies B on their own or gives the B's contacts to a health advisor for assisted notification; (iv') B is indeed infected, moreover, \textit{the infection of either A or B is caused by the pathogen transmission during their contact};  and either (ii) B is successfully reached and notified and (iii) agrees to take a test; or (ii') \textit{by the time of notification, B is already removed by other means} (already got sampled or stopped being infectious without sampling, e.g., died).

The condition (iv') in contrast to (iv) excludes the possibility of notifying partners if there was no pathogen transmission during the contact with them. We additionally simplify the model by allowing to notify only the last such partner. From now on, we will use the term ``partner'' to denote a partner, during a contact with whom a pathogen transmission happened.

Upon notification, if not yet removed,  the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$). Upon sampling, the partner might notify in their turn.

BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate. This creates an observed tip in the transmission tree. 
\end{itemize}

@@

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in~\nameref{S1_Appendix}. The results in section~"\nameref{sec:test}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.

@@

For each model, we generated 100 trees with 500--1\,000 tips. For each tree the parameter values were drawn randomly within the following bounds:
\begin{itemize}
\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
\item removal rate $\psi \in ]0.05, 0.2]$,
\item sampling probability $\rho \in ]0.1, 0.9]$,
\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
\end{itemize}

The common parameters were consistent between models, in the sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models, etc.

@@

Let us assume that there are N cherries in the tree. For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$. We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $2i*$ gets swapped with the tip $2i-1*$ ($1 \leq i \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}. (If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle as in Fig.~\ref{fig:tipswap}). We then calculate the time differences $d_{i'}$ between the tips of these reshuffled cherries. 
The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to month or even a year). Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries. However, if some cherries are ``notified'' their real tip differences should be smaller than reshuffled ones. We therefore output the pvalue calculated by the sign test.

@@

We applied bdpn to each of the trees fixing the sampling probability $\rho$ to its true value (for identifiability). %The results with the becoming-uninfectious rate $\psi$ fixed are similar and given in the Supplementary Material. 
As expected with a maximum-likelihood estimator, the tree likelihoods for the estimated parameter values were higher than or equal to the tree likelihoods for the true parameter values for all the trees in both settings.
